The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-12-31 doc/SA-Hermaphrodites-wDemography-R1.tex
Shortened introduction
Intro magic

\textit{Manuscript elements}: figure~1, figure~2, figure~3, figure~4, table~1, table~2, appendices~A, B, and C, Online Supplementary Material. All figures are to print in color.

@@

\textcolor{blue}{When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other.} Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of such SA genetic polymorphisms, but these models have rarely considered the demographic consequences \textcolor{blue}{ of the maladaptation imposed by coexisting alleles with deleterious fitness effects in each sex.} In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

@@

To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}.

\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and negative correlationpolygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

Alleles with opposing fitness effects through male and female sex functions can also cause SA selection in hermaphrodite populations, where individuals' overall fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. \textcolor{blue}{In obligately outcrossing hermaphrodite populations, SA selection operates similarly as in seperate-sexed ones (dioecious species). However, the possibility of self-fertilization has important consequences for the maintenance of SA genetic variation in hermaphrodites}. Self-fertilization reduces \textcolor{blue}{opportunities for} balanced SA polymorphisms, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014,Glemin2021}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, antagonism between viability and fecundity selection \citep{Glemin2021}, or spatial heterogeneity and complexity of the life-cycle \textcolor{blue}{can all facilitate the maintenance of} SA polymorphisms \citep{Olito-etal-2018,ConnallonSharmaOlito2019, Glemin2021}. Overall, both theoretical predictions and current empirical data suggest that there is ample scope for SA \textcolor{blue}{fitness variation} and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, DelphHerlihy2012, WangBarrett2020, RuzickaESEB2020, RuzickaConnallon2022}.

\textcolor{blue}{While significant efforts have been made to understand when balanced SA polymophisms can be maintained and to detect SA genetic variation from empirical data, much less consideration has been given to the demographic consequences of the maladaptation caused by SA selection}. Most population genetic models, including models of SA selection, model the relative fitness of genotypes in populations of constant (and often infinite) size, and do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). Over the past 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. In particular, theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic changes accompanying allele frequency changes due to selection. Extensions of these theories to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects} all highlight that a fuller understanding of the genetic basis of fitness variation, and the processes that maintain it, requires an explicit treatment of the demographic consequences of evolutionary change.}

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism and population dynamics, but did not analyze the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a,deVriesCaswell2019b} to include both dioecious and hermaphroditic \textcolor{blue}{populations} in order to study when different genetic outcomes of SA selection, including fixation of male- or female-beneficial alleles as well as balanced polymorphisms, are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

>>>>

\textit{Manuscript elements}: figure~1, figure~2, figure~3, figure~4, figure~5, table~1, table~2, Appendices~A, B, C, and D, Online Supplementary Material. All figures are to print in color.

@@

\textcolor{blue}{When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other.} Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of such SA genetic polymorphisms, but these models have rarely considered the demographic consequences \textcolor{blue}{ of coexisting alleles with deleterious fitness effects in each sex.} In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

@@

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends upon the nature of the genetic variation that it harbors \citep{Fisher1930, Lewontin1975}. Yet, species exist in a complex world where the process of adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{CharlesworthHughes2000, ConnallonHall2018}.}

Genetic constraints on adaptation due to conflicting selection have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which shapes a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy, harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

%\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects through male and female sex functions can cause SA selection in hermaphrodite populations, where individuals' fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA \textcolor{blue}{genetic variation} in both dioecious \textcolor{blue}{(separate-sexed)} and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020}.}

% Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

% Alleles with opposing fitness effects through male and female sex functions can also cause SA selection in hermaphrodite populations, where individuals' overall fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. \textcolor{blue}{In obligately outcrossing hermaphrodite populations, SA selection operates similarly as in seperate-sexed ones (dioecious species). However, the possibility of self-fertilization has important consequences for the maintenance of SA genetic variation in hermaphrodites}. Self-fertilization reduces \textcolor{blue}{opportunities for} balanced SA polymorphisms, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014,Glemin2021}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, antagonism between viability and fecundity selection \citep{Glemin2021}, or spatial heterogeneity and complexity of the life-cycle \textcolor{blue}{can all facilitate the maintenance of} SA polymorphisms \citep{Olito-etal-2018,ConnallonSharmaOlito2019, Glemin2021}. Overall, both theoretical predictions and current empirical data suggest that there is ample scope for SA \textcolor{blue}{fitness variation} and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, DelphHerlihy2012, WangBarrett2020, RuzickaESEB2020, RuzickaConnallon2022}.

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms should be maintained and to detect SA genetic variation from empirical data, less attention has been paid to the demographic consequences of SA selection \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and the evolution of quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. Together, these eco-evolutionary models highlight that a fuller understanding of the genetic basis of fitness variation, and the processes maintaining it, requries an integration of genetic and demographic evolutionary processes.}

%a fuller understanding of the genetic basis of fitness variation, and the processes maintaining it, requires an explicit treatment of the demographic processes involved in evolutionary change.

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with SA selection and population dynamics, but did not analyze how demography influenced the maintenance of SA polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a,deVriesCaswell2019b} to \textcolor{blue}{include both dioecious and hermaphroditic populations and jointly analyze the genetic (fixation of male- or female-beneficial alleles, or balanced polymorphism) and demographic (population persistence or extinction) outcomes of SA selection}. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).
